The Gender Pay Gap in the General Social Survey

### 2023-09-22 Paper/paper.Rmd

In this paper we predict individuals' wages using data from the General Social Survey (GSS, 1974-2018). The variable of interest is `realrinc` which varies between 227\$ and 480,145\$\footnote{In constant 1986 USD.}. Hence, we can use regression techniques to predict individuals' wages. Additionally, we can consider whether `realrinc` varies by gender and if so, how much.

@@

The prediction of individuals' wages has been an issue in previous literature. As wages are relevant to companies they want to know how to predict wages [@Chakraborti]. Some papers make use of classification techniques to predict wages [@Chakraborti]. As we have numeric values of real wages and not a categorical variable we do not make use of classification techniques but prefer linear regression techniques such as OLS and Lasso regressions as well as random forests. We prefer these techniques as they handle categorical variables and outliers well, which are features of our data [@Cutler]. Wages were already predicted using random forests by [@Eichinger]. They find that using random forests results in better predictions than using linear models. Additionally, they show that gender barely effects the wage predictions. Likewise, the boosting methods used by [@Chakrabarty2018] in salary level classification problems showed better performance compared to a traditional algorithm such as support vector machines (SVM). Furthermore, @BonaccoltoToepfer1612 suggest the use of a penalization approach for high dimensional problems in his research about gender wage inequality.

@@

This data set still has some missing values for the regressors, e.g., number of children, age, marital status, degree level, and others. Thus, we use the `mice` package with five variables for the imputation process. The features *age, occupational prestige and number of children* are imputed using the predictive mean matching (pmm) method, and we use the polytomous logistic regression method for the factor variables (*education and marital status*). As we use a value of 6 in the `m` parameter of the `mice` function, the imputation process generates 6 different data sets. This parameter refers to the number of imputed data sets. After this, we apply the calculation of the mean and the majority vote to aggregate the data sets into one. Mean estimation is applied for numerical variables and majority voting for categorical variables

@@

As the correlation assesses the linear relationship between two variables and it only uses numerical features, we additionally estimate the Predictive Power Score (PPS). It overcomes these limitations and allows us to assess the predictive power of a regressor on a target variable. This measure is a normalized metric and its values are between 0 (no predictive power) and 1 (perfect predictive power). The PPS is calculated by comparing the performance of a model that uses a regressor `x` to predict target variable `y` versus the performance of a base model (naive). When the target variable is a numerical feature it is common to use a Decision Tree Regressor to predict `y` based on `x`, while in categorical cases a Decision Tree Classifier is used. On the other hand, base models are often defined as mean prediction or random or modal prediction for regression and classification problems, respectively. The PPS results are presented in figure 3, and show that `prestg10`, `wrkstat`, `educcat` and `occrecode` have the highest predictive power of `realrinc` relative to the other features. These results provide some insights in the data exploration process, but we must beware of the limitations of this approach.

@@

After the data preparation and feature engineering processes, we have a final data set with 33,244 observations and 421 features. Due to the large number of interaction variables, we reduce them and use only the interaction between a group of variables (`occrecode`, `educcat` and categorical variable of `prestg10`). After this, we develop the modeling process with 100 variables (numerical and interaction variables). Figure 4 shows the overview of the group of estimated variables that are used in the models.

@@

The best model using the logarithmic transformation has an RMSE of 34,385.47 on the test data set. Thus, the error of this model is greater relative to the best model without the transformation (29,207.55). These results show that in tree-based methods, Gradient Boosting has a dominant performance over Random Forest. In other words, the error of Gradient Boosting is about 8% smaller than the error of Random Forest when the logarithmic transformation is not used. Nevertheless, the errors of these models are larger than those of the linear models.

@@

The best model obtaines an RMSE of 25,856.76 on the test data set. Thus, the error of this model is slightly lower in relation to the linear model that also uses the logarithmic transformation (25,909.82). In other words, it is 0.20% better than the initial linear model. Figure 9 exposes the approximate significance of the smooth terms, and the results show that all the smooth terms used in the model have a p-value of less than 0.01.

>>>>

In this paper we predict individuals' incomes using data from the General Social Survey (GSS, 1974-2018). The variable of interest is `realrinc` which varies between 227\$ and 480,145\$\footnote{In constant 1986 USD.}. Hence, we can employ regression techniques to predict individuals' incomes. Additionally, we can consider whether `realrinc` varies by gender and if so, how much.

@@

The prediction of individuals' incomes has been an issue in previous literature. As wages are relevant to companies they want to know how to predict wages [@Chakraborti]. Some papers make use of classification techniques to predict wages [@Chakraborti]. Since we have numeric values representing real wages and not a categorical variable we do not make use of classification techniques but prefer linear regression techniques such as OLS and Lasso regressions as well as random forests. We prefer these techniques as they handle categorical variables and outliers well, which are features of our data [@Cutler]. Wages were already predicted using random forests by [@Eichinger]. They find that using random forests results in better predictions than using linear models. Additionally, they show that gender barely effects the wage predictions. Likewise, the boosting methods used by [@Chakrabarty2018] in salary level classification problems showed better performance compared to a traditional algorithm such as support vector machines (SVM). Furthermore, @BonaccoltoToepfer1612 suggest the use of a penalization approach for high dimensional problems in his research about gender wage inequality.

@@

This data set still has some missing values for the regressors, e.g., number of children, age, marital status, degree level, and others. Thus, we use the `mice` package with five variables for the imputation process. The features *age, occupational prestige and number of children* are imputed using the predictive mean matching (pmm) method, and we use the polytomous logistic regression method for the factor variables (*education and marital status*). As we use a value of 6 in the `m` parameter of the `mice` function, the imputation process generates six different data sets. This parameter refers to the number of imputed data sets. After this, we apply the calculation of the mean and the majority vote to aggregate the data sets into one. Mean estimation is applied for numerical variables and majority voting for categorical variables

@@

As the correlation assesses the linear relationship between two variables and it only uses numerical features, we additionally estimate the Predictive Power Score (PPS). It overcomes these limitations and allows us to assess the predictive power of a regressor on a target variable. This measure is a normalized metric and its values are between 0 (no predictive power) and 1 (perfect predictive power). The PPS is calculated by comparing the performance of a model that uses a regressor `x` to predict target variable `y` versus the performance of a base model (naive). When the target variable is a numerical feature it is common to use a Decision Tree Regressor to predict `y` based on `x`, while in categorical cases a Decision Tree Classifier is used. On the other hand, base models are often defined as mean, random or modal prediction for regression and classification problems, respectively. The PPS results are presented in figure 3, and show that `prestg10`, `wrkstat`, `educcat` and `occrecode` have the highest predictive power of `realrinc` relative to the other features. These results provide some insights in the data exploration process, but we must beware of the limitations of this approach.

@@

After the data preparation and feature engineering processes, we have a final data set with 33,244 observations and 421 features. Due to the large number of interaction variables, we reduce them and use only the interactions among a selected group of variables (`occrecode`, `educcat` and categorical variable of `prestg10`). After this, we develop the modeling process with 100 variables (numerical and interaction variables). Figure 4 shows the overview of the group of estimated variables that are used in the models.

@@

The best model using the logarithmic transformation has an RMSE of 34,385.47 on the test data set. Thus, the error of this model is greater relative to the best model without the transformation (29,207.55). These results show that in tree-based methods, Gradient Boosting has a superior performance over Random Forest. In other words, the error of Gradient Boosting is about 8% smaller than the error of Random Forest when the logarithmic transformation is not used. Nevertheless, the errors of these models are larger than those of the linear models.

@@

The best model obtaines an RMSE of 25,856.76 on the test data set. Thus, the error of this model is slightly lower in relation to the linear model that also uses the logarithmic transformation (25,909.82). In other words, it performs 0.20% better than the initial linear model. Figure 9 exposes the approximate significance of the smooth terms, and the results show that all the smooth terms used in the model have a p-value of less than 0.01.
